---
title: Data preparation # Also check out line 43
author: Claudius Gräbner-Radkowitsch
date: '2025-05-23'
execute: 
  freeze: auto
format: 
  html: # Settings set in _metadata.yml
    toc: true
  pdf:
    include-in-header: 
      text: |
        \usepackage{lmodern} 
        \usepackage{graphicx}
        \usepackage{hyperref}
        \usepackage{url}                
        \usepackage{xcolor}
        \usepackage{booktabs}
        \usepackage{listings}
        \lstloadlanguages{R}
        
        \definecolor{eufblue}{RGB}{0,57,91}
        \definecolor{eufgrey}{RGB}{111,111,111}
        \definecolor{euflightblue}{RGB}{105,170,205}
        
        \hypersetup{
        pdfauthor={Claudius Graebner-Radkowitsch}
        colorlinks=true,
        linkcolor=euflightblue,
        urlcolor=euflightblue
        }
        \usepackage[includehead,includefoot,top=2cm, bottom=1.5cm]{geometry}
        \usepackage[headsepline, footsepline]{scrlayer-scrpage}
        \pagestyle{scrheadings}
        \clearpairofpagestyles
        \ihead{Tutorial: First steps in R}
        %\chead{Kopfzeile Mitte}
        \ohead{\pagemark} %
        \ifoot{}
        \cfoot{\href{https://euf-datascience-spring24.netlify.app/}{\normalfont\color{eufblue}{Data Science Using R - Spring Semester 2024}}} % Fußzeile Mitte
        \ofoot{} 
        \setkomafont{disposition}{\color{eufblue}\bfseries}
---


# Packages and data used

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(data.table)
library(here)
```

```{r}
#| label: create-data-raw
#| include: false
#| eval: false

data_raw <- WDI::WDI(
  country = c("DE", "AT", "GR", "IT"), 
  start = 1990, 
  indicator = c("unemp"="SL.UEM.TOTL.ZS", 
                "gdp"="NY.GDP.PCAP.PP.KD",
                "share_indus"="NV.IND.TOTL.ZS",
                "co2"="EN.ATM.CO2E.PC")
  ) %>%
  select(-iso2c) %>%
  filter(
    country %in% c("Germany", "Greece"),
    year %in% c(2017, 2018)
  ) %>%
  select(
    all_of(c("country", "year", "unemp", "gdp"))
  )

fwrite(
  x = data_raw, 
  file = "data/wrangling_data_raw.csv")
```

```{r}
#| label: create-data-raw-long
#| include: false
#| eval: false

data_raw <- fread(file = "data/wrangling_data_raw.csv")

data_raw_long <- tidyr::pivot_longer(
  data = data_raw, 
  cols = dplyr::all_of(c("unemp", "gdp")), 
  names_to = "variable", 
  values_to = "value")

fwrite(
  x = data_raw_long, 
  file = "data/wrangling_data_raw_long.csv")
```


```{r}
#| label: create-gdp-join
#| include: false
#| eval: false

gdp_join <- data_raw %>%
  select(-unemp)

fwrite(
  x = gdp_join, 
  file = "data/wrangling_gdp_join.csv")
```


```{r}
#| label: create-gini-join
#| include: false
#| eval: false

gini_join <- fread(file = "data/swiid9_2_summary.csv") %>%
  filter(
    country %in% c("Greece"),
    year %in% c(2015, 2017)
  ) %>%
  select(
    all_of(c("country", "year", "gini_disp"))
  ) %>%
  rename(gini=gini_disp)

fwrite(
  x = gini_join, 
  file = "data/wrangling_gini_join.csv")
```

```{r}
#| label: create-data-final-expl
#| include: false
#| eval: false

data_final_expl <- WDI::WDI(
  country = c("DE", "AT", "GR", "IT"), 
  start = 2000, 
  indicator = c("unemp"="SL.UEM.TOTL.ZS", 
                "gdp"="NY.GDP.PCAP.PP.KD")
  ) %>%
  select(-iso2c)

fwrite(
  x = data_final_expl, 
  file = "data/wrangling_data_final_expl.csv")
```

```{r, include=FALSE}
data_raw <- fread(file = "data/wrangling_data_raw.csv")
data_raw_long <- fread(file = "data/wrangling_data_raw_long.csv")
data_final_expl <- fread(file = "data/wrangling_data_final_expl.csv")
gini_join <- fread(file = "data/wrangling_gini_join.csv")
gdp_join <- fread(file = "data/wrangling_gdp_join.csv")
```

The data sets used in these notes are available from the 
[course homepage](data-examples.zip):

* `wrangling_data_raw.csv` (`data_raw`)
* `wrangling_data_raw_long.csv` (`data_raw_long`)
* `wrangling_data_final_expl.csv` (`data_final_expl`)
* `wrangling_gini_join.csv` (`gini_join`)
* `wrangling_gdp_join.csv` (`gdp_join`)

The brackets show the names of the data sets used below.

# General remarks

* If you imported the data from a file, make sure that the import went as expected
* Start with a data set that is of the type `tibble` (use `tibble::as_tibble()` if necessary)
* Before starting to wrangle, make a note to yourself of how the final data set should look like; 
  * Then think about the different steps you need to take to reach this goal;
  * Each step should only address one single wrangling challenge
* It is often useful to save the wrangling code in one script in which you 
import raw data in the beginning and save tidy data in the end; this allows you 
to keep data wrangling, visualization, and modelling in separate files

# Reshaping data with the tidyr package

At the end of each data wrangling activity, you should have created 
[tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html). 
The R package that contains functions meant to assist you in 
this process is called [tidyr](https://tidyr.tidyverse.org/).
While it contains functions for various purposes, the by far most common
(and, to be honest, to most demanding) task is to reshape data from long to 
wide format and vice versa. This will be the focus of this section. 
Tutorials on the other `tidyr`-functions can be found on the official package
documentation, but these functions are more specific and not as important
in daily practice as those for reshaping data.

## Wide and long format: definition

There is no strict definition for wide and long data. Rather, the two 
should be understood as *relative* descriptions of data, meaning that it is
more straightforward to speak of a data set that is long*er* relative to
another one, rather than *a long* data set per se.

Here is an example for a rather long data set:

```{r, echo=FALSE}
data_raw_long
```

Here, we have one column identifying the variable, the value of which is 
stored in a separate column.
This means that the data is relatively 'long' in the sense of having
many rows. At the same time, it is relatively 'narrow' in the sense of not
having too many columns since the variable identifier is kept in a single 
column.

Contrast this with an example for a rather wide data set, where each variable
has its own column:

```{r, echo=FALSE}
data_raw
```

While the number of columns remains the same, the data set has relatively more
columns as compared to the rows. At the same time, it tends to be shorter in
the sense of having fewer rows.^[If we had a data set with three instead of
two variables, the wide data set would have the same number of rows, but 
more columns, i.e. it would be wider in an absolute sense as well.]

While the long format is often easier to read and preferable when communicating
data to humans, making data tidy often involves the task of making data 
'longer'.

## Transforming long data into wide data

To make data wider we use the function `tidyr::pivor_wider()`.

Assume that we start with our long data set introduced above and that this
data set is bound to the name `data_raw_long`.

```{r}
dplyr::glimpse(data_raw_long)
```

We will now use `tidyr::pivor_wider()` to make this data set wider. The most
important arguments of this function are as follows:^[
  The function allows for much more finetuning. 
  You might read more about its argument in the help page of the function or the 
  [online documentation](https://tidyr.tidyverse.org/reference/pivot_wider.html).]

* `data` is the first argument and refers to the name of the data set to be considered
* `names_from` denotes the column that includes the names of the new columns
* `values_from` denotes the column that includes the values to be allocated in the newly created cells

In the present case, the call would look like the following:

```{r}
data_raw_wide <- tidyr::pivot_wider(
  data = data_raw_long, 
  names_from = "variable", 
  values_from = "value")
data_raw_wide
```

## Transforming wide data into long data

Assume we want to take the data set `data_raw_wide` and re-create the original
long version. To achieve this we can use `tidyr::pivot_longer()`. Again, lets
have a look at the most important arguments:^[
  See the [online documentation](https://tidyr.tidyverse.org/reference/pivot_longer.html) 
  for a more complete description.]
  
* `data` is the first argument and refers to the name of the data set to be considered
* `cols` denotes the columns that should be transformed into the longer format
* `names_to` denotes the column that includes the names of the new columns
* `values_to` denotes the column that includes the values to be allocated in the newly created cells

The arguments `names_to` and `values_to` are not strictly necessary since they
have useful default values, but its usually nicer to be explicit.

When specifying the argument `cols` you have several possibilities. The 
simplest variant is to pass a character vector with the column names. 
But note that you can save a lot of writing by using so called
[selection helpers](https://dplyr.tidyverse.org/reference/dplyr_tidy_select.html), 
a very useful tool we will learn about later.

In our case this amounts to:

```{r}
data_raw_long <- tidyr::pivot_longer(
  data = data_raw_wide, 
  cols = c("unemp", "gdp"), 
  names_to = "indicator", 
  values_to = "values")
data_raw_long
```


# Chaining wrangling tasks using pipes

Pipes are provided via the package `magrittr`, which is loaded automatically if
you attach packages such as `tidyr` or `dplyr`. Pipes are short keywords that
facilitate the development of very readable and transparent data wrangling code.

While there are many 
[different pipes](https://magrittr.tidyverse.org/reference/index.html), 
the one we will use extensively is `%>%`.
It is always used at the end of a line, and it basically 'throws' the result
of this line into the next line of code. In this line, you can refer to the 
intermediate result via `.`, or it is used implicitly as the first argument
to the function you use.

In other words, `x %>% f(y)` (or `x %>% f(., y)`) is equivalent to `f(x, y)`.

But lets look at an example! Assume we start with the following data set:

```{r, echo=TRUE}
pipe_data_raw <- data_raw %>%
  select(country, year, gdp, unemp)
pipe_data_raw
```

And what we want is this:

```{r, echo=FALSE}
pipe_data_raw %>%
  pivot_longer(cols = c("gdp", "unemp")) %>%
  pivot_wider(names_from = "year", values_from = "value")
```

We can do this by first making the data longer, and then wider. We could do 
this explicitly:

```{r}
pipe_data_1 <- pivot_longer(
  data = pipe_data_raw, 
  cols = all_of(c("gdp", "unemp")))

pipe_data_2 <- pivot_wider(
  data = pipe_data_1,
  names_from = "year", 
  values_from = "value") 
```


But we can also write this code more concisely using the pipe:

```{r}
pipe_data_final <- pivot_longer(
  data = pipe_data_raw, 
  cols = all_of(c("gdp", "unemp"))) %>%
  pivot_wider(
  data = .,
  names_from = "year", 
  values_from = "value") 
```


Or, since the pipe carries the intermediate result implicitly as the first
argument to the function on the next line we can space the `data = .`:

```{r}
pipe_data_final <- pivot_longer(
  data = pipe_data_raw, 
  cols = all_of(c("gdp", "unemp"))) %>%
  pivot_wider(
  names_from = "year", 
  values_from = "value") 
```

The `%>%`-pipe allows you to write very readable code, so make sure you use it
often. But for code development it might be nevertheless helpful to write the
intermediate steps explicitly.

# Manipulating data with the dplyr package

The package `tidyr` is for turning messy data sets into tidy data sets. 
Another important part of data preparation is to manipulate an existing data
set. To do this, the package [dplyr](https://dplyr.tidyverse.org/) contains 
many useful functions. You will regularly use functions from both the `tidyr` and
`dplyr` package within your script for data preparation: often, you start with
a messy data set, make it tidy using functions from `tidyr` and then use
functions from `dplyr` to manipulate variables until your data set contains all 
the information you want. Thus, in the following we cover the most frequently
used `dplyr`-functions. A complete documentation of the whole package can be
found on the official [dplyr webpage](https://dplyr.tidyverse.org/).

## Creating or manipulating variables

The function `dplyr::mutate()` is used both for manipulating existing 
columns as well as creating new columns. In the first case the name of the 
column that the result of `dplyr::mutate()` is written into already exists,
in the second case we just use a new name.

Consider the following data set with the unemployment rate as an example:

```{r, include=FALSE}
data_unemp <- data_raw %>%
  select(c("country", "year", "unemp")) %>%
  pivot_wider(names_from = "country", values_from = "unemp") 
```


```{r}
data_unemp
```
Assume we want to express the percentage values via decimal numbers and, to
this end, divide the values in the columns `Germany` and `Greece` by 100. We
can use `dplyr::mutate()` to achieve this:

```{r}
data_unemp %>%
  dplyr::mutate(
    Germany = Germany/100,
    Greece = Greece/100
  )
```

But we could use basically the same code to create a new column. Assume, for
instance, we want a new column containing the difference between the 
unemployment rates:

```{r}
data_unemp %>%
  dplyr::mutate(
    Difference = Greece - Germany
  )
```

The only difference here was that the left-hand-side name of the column to 
be manipulated did not exist before!

## Filtering rows

The function `dplyr::filter()` can be used to filter rows according to 
certain conditions. The conditions must evaluate for each cell entry to 
either `TRUE` or `FALSE`, and only those rows for which they evaluate to `TRUE`
remain in the data set. Often, the conditions are specified via 
logical operators, which were already covered in the tutorial on 
[vector types](/tutorial/object-types-vec/).

As always, the first argument to `dplyr::filter()` is `data`, i.e. the data
set on which you want to operate. Then follow an arbitrary number of logical
conditions on the different columns of the data set on question. 

Assume we want to take the previously defined data set `data_raw_long`

```{r}
data_raw_long
```

and only want to keep data on GDP:

```{r}
data_raw_long %>%
  dplyr::filter(indicator=="gdp")
```

You may also combine more than one condition in one call to `dplyr::filter()`.
If you also want to filter by values and only keep those rows where the value
is below 50.000:

```{r}
data_raw_long %>%
  dplyr::filter(
    indicator=="gdp",
    values < 50000)
```

## Selecting columns 

When you only want to keep certain *columns* we speak of selecting (rather than
filtering) columns. This is done - surprise - via the function ´dplyr::select()`.

There are different ways for selecting columns. In any case, the first argument
is, again, `data`, i.e. the data set considered. In the present case, we will
refer to `data_raw`:

```{r}
data_raw
```

Then we can now select columns using one of the following two options.
First, you may refer to columns via their *name*:

```{r}
data_raw %>%
  dplyr::select(country, year, unemp)
```

But this is often error-prone. Thus, it is usually better to refer to the columns via 
[selection helpers](https://dplyr.tidyverse.org/reference/dplyr_tidy_select.html),
which is also the most flexible version. While we will learn about more selection 
helpers later, here we will mainly use `dplyr::all_of()`, which accepts a character
vector of column names:

```{r}
data_raw %>%
  dplyr::select(dplyr::all_of(c("country", "year", "gdp")))
```

> **Caution:** Do not forget the `c()`! Otherwise:

```{r, error=TRUE}
data_raw %>%
  dplyr::select(dplyr::all_of("country", "year", "gdp"))
```

> It is also possible to define the column vector first:

```{r}
cols2keep <- c("country", "year", "gdp")
data_raw %>%
  dplyr::select(dplyr::all_of(cols2keep))
```

> Selection helpers allow you to specify the columns to be selected more
generally. For instance, `dplyr::ends_with()` allows you to select all colums
that end with a certain pattern:

```{r}
data_raw %>%
  dplyr::select(dplyr::ends_with("p"))
```

In any case, you can also specify the columns you want to *drop*. To this end, 
just add a `-` in front of the selection command:

```{r}
data_raw %>%
  dplyr::select(-unemp, -gdp)
```

## Merging data sets

Often you need to obtain data from different sources. To merge all your data
in one single data set, you need to use one of the `*_join()` functions of the
`dplyr`-package. These functions all merge two data sets, but the way they do
it is different. Below we illustrate the most common joins (so called 
[mutating joins](https://dplyr.tidyverse.org/reference/mutate-joins.html)).^[
  The other join types are 
  [filtering joins](https://dplyr.tidyverse.org/reference/filter-joins.html) and 
  [nest joins](https://dplyr.tidyverse.org/reference/nest_join.html). You find 
  more information in the web, and more details on the underlying theory in 
  [chapter 13 of R4DS](https://r4ds.had.co.nz/relational-data.html).]

As a guiding example we use the following two data sets:

First, data on income inequality from the 
[SWIID data base](https://fsolt.org/swiid/):

```{r}
gini_join
```

Second, data on GDP per capita from the 
[World Bank](https://data.worldbank.org/indicator/NY.GDP.PCAP.PP.KD):

```{r}
gdp_join
```

We will consider the behavior of the following four functions:

* `dplyr::left_join()`
* `dplyr::right_join()`
* `dplyr::full_join()`
* `dplyr::inner_join()`

All of them accept the following arguments:

* `x` and `y`: the two data sets to be merged
* `by`: a vector or a named vector indicating on which columns the data sets 
should be merged

Its easier to understand their behavior if you contrast them directly with
each other. First, `dplyr::left_join()` joins the data sets on those 
columns mentioned in `by`, but only keeps those rows for which `x` contains
an observation:

```{r}
dplyr::left_join(x = gdp_join, y = gini_join, by = c("country", "year"))
```

This might introduce `NA`s into the columns of `y`, but not of `x`. It is the
other way around for `dplyr::right_join()`: it only keeps those rows for 
which `y` contains an observation:

```{r}
dplyr::right_join(x = gdp_join, y = gini_join, by = c("country", "year"))
```

`dplyr::inner_join()` is the most restrictive option, keeping only those rows
for which both `x` and `y` contain an observation (i.e. it never introduces
`NA`s):

```{r}
dplyr::inner_join(x = gdp_join, y = gini_join, by = c("country", "year"))
```

Finally, `dplyr::full_join()` contains all rows that occur at least in `x` *or*
`y`, i.e. it might introduce `NA`s in both the columns of `x` and `y`:

```{r}
dplyr::full_join(x = gdp_join, y = gini_join, by = c("country", "year"))
```


Two final remarks: first, the types of the columns on which you merge the 
data sets must be equal, otherwise R throws an error:

```{r, error=TRUE}
gini_join <- dplyr::mutate(gini_join, year=as.character(year))
dplyr::left_join(x = gdp_join, y = gini_join, by = c("country", "year"))
```

Just enforce the correct data type before merging:

```{r, error=TRUE}
gini_join %>% 
  dplyr::mutate(year=as.integer(year)) %>%
  dplyr::left_join(x = gdp_join, y = ., by = c("country", "year"))
```

Second, you can also merge on columns with different names by passing named
vectors to `by`:

```{r}
gini_join <- gini_join %>%
  mutate(Year=as.double(year)) %>%
  select(-year)
gini_join
```

Then this does not work any more:

```{r, error=TRUE}
dplyr::left_join(
  x = gdp_join, y = gini_join, 
  by = c("country", "year"))
```

But the named vector fixes it:

```{r}
dplyr::left_join(
  x = gdp_join, y = gini_join, 
  by = c("country", "year"="Year"))
```

## Grouping and summarising data

The final challenge we consider involves the application of two functions
(at least in most cases): `dplyr::group_by()` and `dplyr::summarize()`.

`dplyr::group_by()` is usually used within pipes and groups a data set 
according to an arbitrary number of variables, each of which must refer to 
one (and only one) column. It produces a grouped data set:

```{r}
data_raw_grouped <- data_raw %>%
  dplyr::group_by(country)
data_raw_grouped
```

As you can see, the data set is now grouped by the variable `country`. We can
specify the grouping variables the same way we selected columns in the context
of `dplyr::select()` (see above).

Grouped data sets are usually not interesting in itself. You can ungroup them
via `dplyr::ungroup()`:

```{r}
data_raw_grouped %>%
  dplyr::ungroup()
```

They are most useful if used in conjunction with `dplyr::summarise()`, which
summarizes variables. While it can be used without
`dplyr::group_by()`, it is most useful if it is applied to grouped data sets:
then it computes summary statistics for each group.

```{r}
data_raw %>%
  summarise(
    avg_gdp=mean(gdp)
  )
```

```{r}
data_raw_grouped %>%
  summarise(
    avg_gdp=mean(gdp)
  )
```

You can also summarized more than one column:

```{r}
data_raw_grouped %>%
  summarise(
    avg_gdp=mean(gdp),
    median_unemp=median(unemp)
  )
```

Note that `dplyr::summarise()` drops all columns that it is not asked to 
compute summary statistics for, except potential grouping variables. There 
are also some advanced features of the functions, which are explained in
the official [documentation](https://dplyr.tidyverse.org/reference/summarise.html).

# A final example

Thanks to the pipes it is easy to chain the many different wrangling steps into
one function call. But in practice it is very important that you (1) inspect
your raw data very clearly, (2) write down the desired end product, and then (3)
think about the single steps required to reach the desired outcome. Each step 
should address one (and only one) wrangling challenge.

To illustrate this, assume we start start with this raw data:

```{r}
str(data_final_expl)
```

What we want to do is to compute the difference in the country averages of the 
variables for the time periods 2005-2007 and 2010-2013. This would look like 
this:^[We have not yet covered the function `ifelse()`. It contains a logical
test as a first argument, and then two further arguments: one return value for
the case in which the test returns `TRUE`, and one for which the test returns
`FALSE`.]

```{r, include=FALSE}
data_final_expl %>%
  dplyr::mutate(
    period = ifelse(
      test = year %in% 2005:2007, 
      yes = "Early", 
      no =  ifelse(
        test = year %in% 2010:2013, 
        yes = "Late", 
        no = NA))
  ) %>%
  dplyr::filter(!is.na(period)) %>%
  group_by(country, period) %>%
  summarise(
    avg_unemp = mean(unemp),
    avg_gdp = mean(gdp), 
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(c("avg_unemp", "avg_gdp")), 
    names_to = "indicator", 
    values_to = "values") %>%
  tidyr::pivot_wider(
    names_from = "period", 
    values_from = "values") %>%
  dplyr::mutate(
    Difference = Late - Early
  ) %>%
  dplyr::select(-Early, -Late) %>%
  tidyr::pivot_wider(
    names_from = "indicator", 
    values_from = "Difference")
```

To achieve this, we need to chain a number of wrangling challenges introduced 
above:

```{r}
data_final_expl %>%
  dplyr::mutate(
    period = ifelse(
      test = year %in% 2005:2007, 
      yes = "Early", 
      no =  ifelse(
        test = year %in% 2010:2013, 
        yes = "Late", 
        no = NA))
  ) %>%
  dplyr::filter(!is.na(period)) %>%
  group_by(country, period) %>%
  summarise(
    avg_unemp = mean(unemp),
    avg_gdp = mean(gdp), 
    .groups = "drop"
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(c("avg_unemp", "avg_gdp")), 
    names_to = "indicator", 
    values_to = "values") %>%
  tidyr::pivot_wider(
    names_from = "period", 
    values_from = "values") %>%
  dplyr::mutate(
    Difference = Late - Early
  ) %>%
  dplyr::select(-Early, -Late) %>%
  tidyr::pivot_wider(
    names_from = "indicator", 
    values_from = "Difference")
```

If you have trouble understanding the many steps, redo the computations 
yourself and always check what happens in the single steps. It is not a good 
idea to write such a long chain in one working step, but rather to make sure 
that you always understand what happens in any single step, and then expand 
the chain one by one.
