---
title: "🗓️ Session 4: Object types in R"
date: "2025-04-04"
date-modified: last-modified
---

In the first part of this session you 
learn some additional details about R packages and about the most important and most fundamental object types in R, such as
decimal numbers or words. While this might look a bit boring at first, understanding
these basic types is fundamental for all the more advanced (and exiting) stuff
in the future! 

In the second part of this session, you are then introduced to 
two more advanced object types in R. 
The two object types covered, factors and
data frames, are advanced in the sense that they can be thought of extensions
of some of the basic object types you encountered before: factors and special
kinds of integers, and data frames are special kinds of lists. 

## 👨‍🏫 Lecture Slides 

For the slides of the first part of this session (*Packages and basic object types*)
on click on the slide area below or click 
[here](/content/material/slides/E4a-Objects.pdf) to download the slides.


```{=html}
<div class="embed-container">
  <iframe 
  src="/content/material/slides/E4a-Objects.pdf" 
  style="border: 0.5px" 
  width="80%"
  height="80%"
  ></iframe>
</div>
```

::: {.callout-tip collapse="true"}
## Solutions to the intermediate exercises of part 1
<script src="https://gist.github.com/graebnerc/de4c233f8901905a5d4c15a586547801.js"></script>
:::

For the slides of the second part of this session (*Advanced object types*) 
on click on the slide area below or click 
[here](/content/material/slides/E4b-AdvObjects.pdf) to download the slides.

```{=html}
<div class="embed-container">
  <iframe 
  src="/content/material/slides/E4b-AdvObjects.pdf" 
  style="border: 0.5px" 
  width="80%"
  height="80%"
  ></iframe>
</div>
```


::: {.callout-tip collapse="true"}
## Notes and solutions for the second part of this session
<script src="https://gist.github.com/graebnerc/daef5dee107ee89ca9cd618fa0dde96d.js"></script>
:::


## 🎥 Lecture videos 

All the videos for the first part on basic object types
are available via this [playlist](https://www.youtube.com/playlist?list=PLZDawQMrG1RJ1oZwuZtiG0fOWSOgyZhYC).

::: {.callout-note collapse="true"}
## Expand to access the videos directly

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/2pTzLhBfjRI?si=CM9MBNOOuvAQ0xGa" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</div>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/8A5x-uENcrQ?si=wAaxGUzltTk5rpDu" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/ETZgjjWbUEM?si=awPGu3J-q5Rosq8v" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/2sTn2nBypEE?si=pJzRnVSf0kAUqcWT" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/pcwTeS3H39w?si=4xSvvkZOqDiY7XpZ" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/qpeRL-Pm0C4?si=SKBEYVM0fIcMaHsN" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
```
:::


All the videos for the second part on advanced object types 
are available via this [playlist](https://www.youtube.com/playlist?list=PLZDawQMrG1RK0Ia3Y3AFqf_FP59GJnIP7).

::: {.callout-note collapse="true"}
## Expand to access the videos directly

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/y-2i8MmT7y8?si=zqoHXyn0R43wDgtD" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</div>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/0-oFA2V6b-Q?si=McN3QrmKbZ_IjWwK" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</div>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/yYR3C77UQC0?si=KPMM21mc-w6k862A" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/GrEdx3T4Ylk?si=eiDv6FSQSoipMPE7" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/TnFbHp7MFRk?si=qcff5TBtPmX3hJlU" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
```
:::

## 📚 Mandatory Reading

Read the following tutorials for the first part on basic object types:

- [Fundamental object types in R I: Functions](/content/tutorials/obj-types-i-functions/index.qmd)
- [Fundamental object types in R II: Vectors](/content/tutorials/obj-types-ii-vectors/index.qmd)

The topics of the second part on advanced object types are covered here:

- [Fundamental object types in R III: Factors and data frames](/content/tutorials/obj-types-iii-adv/index.qmd)

### 🏆 Further readings

*I suggest you read these references after you learned about data wrangling techniques in session 10.*

- Sections 1-3 in [Chapter 12](https://r4ds.hadley.nz/logicals) of @r4ds.
- [Chapter 13](https://r4ds.hadley.nz/numbers) in @r4ds.
- [Chapter 14](https://r4ds.hadley.nz/strings) in @r4ds.
- [Chapter 16](https://r4ds.hadley.nz/factors) in @r4ds. 

## ✍️ Coursework

- Do the `ObjectTypes1` exercises of the package `DataScienceExercises` 
for the first part on basic object types:

::: {.callout-tip collapse="true"}
## Quick code for starting the exercises
```{r}
#| eval: false
learnr::run_tutorial(
  name = "ObjectTypes1", 
  package = "DataScienceExercises", 
  shiny_args=list("launch.browser"=TRUE))
```
:::

- For the second part do the `ObjectTypes2` exercises of the package `DataScienceExercises` 

::: {.callout-tip collapse="true"}
## Quick code for starting the exercises
```{r}
#| eval: false
learnr::run_tutorial(
  name = "ObjectTypes2", 
  package = "DataScienceExercises", 
  shiny_args=list("launch.browser"=TRUE))
```
:::

- If you have questions or problems, please post them in the Moodle forum
