---
title: "🗓️ Session 12: Linear regression in R"
date: "2025-06-06"
date-modified: last-modified
---

This session focuses on the practical implementation of linear regression models.
This way it complements the more theoretical lecture. More precisely, we focus
on the following topics:

- Implement simple and multiple linear regression in R
- Calculate and interpret $R^2$ manually and using `R` functions
- Understand and demonstrate omitted variable bias
- Perform data transformations to linearize relationships
- Create effective visualizations using `geom_smooth()`
- Conduct basic model diagnostics


## 📚 Mandatory Reading

- The tutorial [Linear Regression in R](/content/tutorials/linear-regression/index.qmd) 
  - [Data sets used](/content/tutorials/linear-regression/RegressionDataSets.zip)

## Further Reading

- [Chapter 5](https://moderndive.com/v2/regression.html) and [Chapter 6](https://moderndive.com/v2/multiple-regression.html) in @moderndive

## ✍️ Coursework

- Do the exercises `LinearRegression1` and `LinearRegression2` from the `DataScienceExercises` package

::: {.callout-tip collapse="true"}
## Quick code for starting the exercises I
```{r}
#| eval: false
learnr::run_tutorial(
  name = "LinearRegression1", 
  package = "DataScienceExercises", 
  shiny_args=list("launch.browser"=TRUE))
```
:::

::: {.callout-tip collapse="true"}
## Quick code for starting the exercises II
```{r}
#| eval: false
learnr::run_tutorial(
  name = "LinearRegression2", 
  package = "DataScienceExercises", 
  shiny_args=list("launch.browser"=TRUE))
```
:::


