---
title: "Exercises for Recap Session 1"
subtitle: "Possible solutions"
date: "2024-04-11"
date-modified: last-modified
format: 
    html: 
        highlight-style: github
    pdf: 
        toc: false
---

# Exercise 1: Basic object types I

1. Create a vector containing the numbers `2`, `5`, `2.4` and `11`.

```{r}
ex1_vec <- c(2, 5, 2.4, 11)
```

2. Replace the second element with `5.9`.

```{r}
ex1_vec[2] <- 5.9
ex1_vec
```

3. Add the elements `3` and `1` to the beginning, and the elements 
`"8.0"` and `"9.2"` to the end of the vector.

```{r}
va_1 <- c(3, 1)
va_2 <- c("8.0", "9.2")
ex1_vec_extended <- c(va_1, ex1_vec, va_2)
ex1_vec_extended
```

4. Create a vector with the numbers from -8 to 9 (step size: 0.5)

```{r}
ex1_vec_4 <- seq(-8, 9, by = 0.5)
ex1_vec_4
```

5. Compute the square root of each element of the first vector 
using vectorisation.

```{r}
sqrt(ex1_vec_4)
```

6. Create a character vector containing then strings `"Number_1"` to 
`"Number_5"`. Use suitable helper functions to create this vector quickly.

```{r}
ex1_char_vec <- paste0("Number_", seq(1, 5))
ex1_char_vec
```

# Exercise 2: Basic object types II

Consider the following vector:

```{r}
ex_2_vec <- c(1.9, "2", FALSE)
```

1. What is the type of this vector? Why?

```{r}
typeof(ex_2_vec)
```

Atomic vectors only contain objects of the same type, and there is a 
hierarchy. Elements that themselves are of a type lower in the hierarchy 
are coerced to the same type as the object highest in the hierarchy. The
hierarchy is as as follows:

1. `character`
2. `double`
3. `integer`
4. `logical`

Therefore, the type of `ex_2_vec` is `character`. The underlying reason is that
you can, for instance, always transform a `double` value into a `character` 
but not vice versa.

2. What happens if you coerce this vector into type integer? Why?

```{r}
as.integer(ex_2_vec)
```

Because `integer` is lower in the hierarchy than `character`, the transformation
is not straightforward. By coincidence, the first two elements can actually be
coerced into integers (albeit maybe not with the expected result), but there is 
no way you can transform the logical value `FALSE` into an integer, which is
why a missing value is produced.

3. What does `sum(is.na(x))` tell you about a vector `x`? What is happening here?

```{r}
x <- c(1,2,3,NA,NA,8)
```
First, `is.na(x)` creates a vector with logical values indicating whether a value of 
the original vector is missing (i.e. `NA`):

```{r}
is.na(x)
```

Then, `sum()` computes the sum over this vecor of boolean values:

```{r}
sum(is.na(x))
```

Here, `TRUE` counts as one and `FALSE` as zero, so `sum()` gives the number of 
cases in which `is.na(x)` has evaluated to `TRUE`:


4. Is it a good idea to use `as.integer()` on double characters to round them to 
the next integer? Why (not)? What other ways are there to do the rounding?

No, because `as.integer()` is not acutally rounding numbers (as, for example, `as.integer(2.1)`
would make you think), but only removing the decimal part of the number:

```{r}
as.integer(2.9) # you might expect 2...
```

Better use `round()`:

```{r}
round(2.9)
```



# Exercise 3: Define a function

Create functions that take a vector as input and returns:

1. The last value.

```{r}
get_last_val <- function(x){
  last_val <- x[length(x)]
  return(last_val)
}
```


2. Every element except the last value and any missing values.

```{r}
get_beginning <- function(x){
  beginning <- x[-length(x)] # Removes last value
  na_positions <- which(is.na(beginning)) # Get positions of NA values
  beginning_nonas <- beginning[-na_positions] # Removes these values
  return(beginning_nonas)
} 
```

3. Only even numbers.
    
> Hint: Use the operation `x %% y` to get the remainder from diving `x` by `y`, 
the so called 'modulo y'. For even numbers, the modulo 2 is zero.

```{r}
get_even <- function(x){
  modulo_2s <- x%%2 # Module 2 is zero for even numbers only
  even_nbs <- x[modulo_2s==0] # Keep only those for which modulo 2 is zero
  na_positions <- which(is.na(even_nbs)) # Get positions of NA values
  even_nbs_nonas <- even_nbs[-na_positions] # Removes these values
  return(even_nbs_nonas)
}
```


Apply your function to the following example vector:

```{r}
ex_3_vec <- c(1, -8, 99, 3, NA, 4, -0.5, 50)
```

```{r}
get_last_val(ex_3_vec)
get_beginning(ex_3_vec)
get_even(ex_3_vec)
```


# Exercise 4: Lists

1. Create a list that contains three elements called `'a'`, `'b'` and `'c'`.
The first element should correspond to a double vector with the elements
`1.5`, `-2.9` and `99`.
The second element should correspond to a character vector with the elments
`'Hello'`, `'3'`, and `'EUF'`.
The third element should contain three times the entry `FALSE`.

```{r}
ex_4_list <- list(
  'a' = c(1.5, -2.9, 99),
  'b' = c('Hello', "'3'", 'EUF'),
  'c' = rep(FALSE, 3)
)
```


2. Transform this list into a `data.frame` and a `tibble`. Then apply `str()` to
get information about the respective structure. How do the results differ?

```{r}
ex_4_df <- as.data.frame(ex_4_list)
ex_4_tb <- tibble::as_tibble(ex_4_list)
str(ex_4_list)
str(ex_4_df)
str(ex_4_tb)
```

`str()` only differs with regard to the first line describing the type.

# Exercise 5: Data frames and the study semester distribution at EUF

The package `DataScienceExercises` contains a data set called `EUFstudentsemesters`,
which contains information about the distribution of study semesters of 
enrolled students at the EUF in 2021. You can shortcut the data set as follows:

```{r}
euf_semesters <- DataScienceExercises::EUFstudentsemesters
```

1. What happens if you extract the column with study semesters as a vector
and transform it into a `double`? 

```{r}
unique(euf_semesters[["Semester"]])
semesters <- as.double(euf_semesters[["Semester"]])
unique(semesters)
```

We see that the previous entry `"9 or higher"` has been transformed into `NA`.

2. What is the average study semester of those students being in their 8th or 
earlier semester?

```{r}
mean(semesters, na.rm = TRUE)
```


3. How many students are in their 9th or higher study semester?

```{r}
sum(euf_semesters$Semester=="9 or higher")
```


4. What does `typeof(euf_semesters)` return and why?

```{r}
typeof(euf_semesters)
```

It returns `list`, because while `euf_semesters` is a `tibble`, `typeof()`
always gives the underlying basic object type. For `tibble`s, this is `list`.
