---
title: "🗓️ Session 10: Data preparation"
date: "2025-05-23"
date-modified: last-modified
---

In this session you learn how to turn your raw data into a state such that you
can work with it. Luckily, there is one particular form for our data that 
represents the common starting point for all further operations, such as 
visualization or modelling. This form is called *tidy data*. And the goal of 
this session is to equip you with the tools that you need to turn the often
messy raw data into tidy data. These skills are important because they make 
you independent: you will be able to prepare any data you find or create 
yourself such that you can further process it, and you will not rely on others
to provide you data in a particular form.


## 👨‍🏫 Lecture Slides 

Either click on the slide area below or click 
[here](/content/material/slides/E10-DataPreparation.pdf) to download the slides.


```{=html}
<div class="embed-container">
  <iframe 
  src="/content/material/slides/E10-DataPreparation.pdf" 
  style="border: 0.5px" 
  width="80%"
  height="80%"
  ></iframe>
</div>
```

::: {.callout-tip collapse="true"}
## Data and solutions to the intermediate exercises

- [Data used in the video](DataWranglingVideo.zip)
- [Data for exercises](DataPrepExData.zip)

<script src="https://gist.github.com/graebnerc/029ca3b65025478f19f75ebfda2c1292.js"></script>
:::


## 🎥 Lecture videos 

All the videos are available via this [playlist](https://www.youtube.com/playlist?list=PLZDawQMrG1RIzp07hiVKmu6wUs8tqdKod).

::: {.callout-note collapse="true"}
## Expand to access the videos directly

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/0MMPyRCEufg?si=wODXoBKkK_7uqy0n" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</div>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/OdCcBWHJKvU?si=Q9doQYZFnodFnI6E" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</div>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/9L0xAgymcKo?si=lSlKxnPq_OsCtxrg" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</div>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/7Sp1U0CLlts?si=ffij8rCZLlXMkqBq" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</div>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/khegEBj6Bqk?si=9RnXuia3lolgYPJF" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</div>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/K4MPQkRZUhw?si=aneKHgSZnH8Y2ZUb" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</div>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/HGkrF-_tvog?si=dpqO9urb1uRIdhwy" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</div>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/1fLZoRsSi-M?si=RBLqN0weVmH_g7kX" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</div>
```

```{=html}
<div class="embed-container">
<iframe width="370" height="208" src="https://www.youtube.com/embed/gaA2j6CLVRQ?si=QHvM2x337ZYk_PnO" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" allowfullscreen></iframe>
</div>
```
:::

::: {.callout-tip collapse="true"}
## Solutions to the intermediate exercises
<script src=""></script>
:::

## 📚 Mandatory Reading

- The tutorial [Data preparation](/content/tutorials/data-wrangling/index.qmd) 
- [Chapter 5](https://r4ds.hadley.nz/data-tidy.html) in @r4ds.

## Further Reading

- The help page for the 
[selection helpers](https://dplyr.tidyverse.org/reference/select.html), 
which facilitate the selection of particular columns.
- [Chapter 13](https://r4ds.hadley.nz/numbers) in @r4ds.
- [Chapter 14](https://r4ds.hadley.nz/strings) in @r4ds.
- @tidydata on the concept of 'tidy data' (see the [Github repo](https://github.com/hadley/tidy-data) for reproduction of the paper)

## ✍️ Coursework

- Do the exercises `Wrangling1` from the `DataScienceExercises` package

::: {.callout-tip collapse="true"}
## Quick code for starting the exercises
```{r}
#| eval: false
learnr::run_tutorial(
  name = "Wrangling1", 
  package = "DataScienceExercises", 
  shiny_args=list("launch.browser"=TRUE))
```
:::
- Download data about the CO2 emissions for some countries of your choice from
the World Bank [website](https://data.worldbank.org/indicator/EN.ATM.CO2E.PC)
for the years 2000 to 2020. Set up an R project, save the data, import it, and
make a line graph.
    - [ZIP file](/content/material/session10-solution.zip) with a possible solution^[You can ignore the `make_co2_data.R` for now and only look at `make_co2_plot.R`.]
- If you want more exercises on the challenge of making data longer/wider, you can do the exercises `Wrangling2` from the `DataScienceExercises` package

::: {.callout-tip collapse="true"}
## Quick code for starting the exercises
```{r}
#| eval: false
learnr::run_tutorial(
  name = "Wrangling2", 
  package = "DataScienceExercises", 
  shiny_args=list("launch.browser"=TRUE))
```
:::
